---
title: "tidy data"
author: "Luis Fernando Delgado"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
rm(list = ls())

source("https://raw.githubusercontent.com/Cassava2050/PPD/main/utilities_tidy.R")
```


## Load the files to check

```{r}
local_file <- "yes" #

if (local_file == "yes") {
  folder <- here::here("data//")  
  file <- "phenotype1.csv"
  skip_col <- 3 # double check the number of col skipped
  trial_interest = "BCPRC"
  year_interest <- 2023
}

# 1) load the data
sel_data <- read_cassavabase(phenotypeFile = paste0(folder, file))
```

## Add the other data set
```{r}
if (local_file == "yes") {
  folder <- here::here("data//")  
  file <- "phenotype.csv"
  skip_col <- 3 # double check the number of col skipped
  trial_interest = "BCPRC"
  year_interest <- 2023
}

# 1) load the data
sel_data1 <- read_cassavabase(phenotypeFile = paste0(folder, file))
```

# bind both data sets
```{r}
sel_data <- sel_data %>% bind_rows(sel_data1)
```


# ---- Change columns into standar names ----

```{r}
sel_data_kp <- change_colname(sel_data, NA)
```


## change the column class

```{r}
obs_col <- c(
  names(sel_data_kp)[str_detect(names(sel_data_kp), "obs_")],
  "use_rep_number", "blockNumber",
  "use_plot_number", "use_plot_width",
  "use_plot_length"
)
sel_data_kp %<>%
  mutate(across(all_of(obs_col), as.numeric))

# remove - , replace by _
names(sel_data_kp) = gsub("-", "_", names(sel_data_kp))
```

## Duplications in row and cols

```{r}
duplicated_plot <- row_col_dup(sel_data_kp)
```

## Plot trial layout
```{r}
trial_layout(sel_data_kp)
```


## Check the clone name
```{r}
cloneName_new_old <- check_clone_name(
  clone_list = sel_data_kp$use_accession_name,
  new_names = NA,
  add_check = NULL
)

trial_standard <- sel_data_kp %>%
  left_join(cloneName_new_old,
            by = c("use_accession_name" = "accession_name_ori")
  ) %>%
  select(-use_accession_name) %>%
  rename(use_accession_name = use_accession_name.y)

```

## Add GIS data

```{r}
trial_standard <- add_GIS(trial_standard)
```

# extract checks

```{r}
accession_rep_ct <- trial_standard %>%
  count(use_trial_name, use_accession_name, use_rep_number)  %>%
  arrange(use_trial_name) %>%
  filter(n>1)
accession_rep_ct 


conducted_trials <- 
  trial_standard %>% group_by(use_trial_name, use_plant_date,use_harvest_date, use_location) %>% 
  summarise(n_gen = n_distinct(use_accession_name)) %>% 
  mutate(harvesting_time = 
           interval(ymd(use_plant_date), ymd(use_harvest_date)) %>% as.period,
         harvesting_time = paste0(harvesting_time@month, "month ", harvesting_time@day, "day")) %>% 
  ungroup()

conducted_trials

conducted_trials %>% relocate(harvesting_time, .after = use_harvest_date) %>% 
  write.table("clipboard", sep="\t", col.names = T, row.names = F)
```


## plot plant number

```{r}
plants_plot <- trial_standard %>%
  group_by(use_trial_name) %>%
  count(obs_planted_number_plot) 
plants_plot

```


## Frequency harvest plant number
```{r}

plants_harvested <- trial_standard %>%
  group_by(use_trial_name) %>%
  count(obs_harvest_number) %>% arrange(desc(obs_harvest_number))
```


# Planted and harvested
```{r}
plants_plot %>% select(-n) %>% 
  left_join(plants_harvested %>% 
              summarise(harvested_plants = max(obs_harvest_number, na.rm = TRUE)), 
            by = "use_trial_name") %>% 
  write.table("clipboard", sep="\t", col.names = T, row.names = F)
```

# Harvested plants

```{r}
plants_to_harvest <- plants_harvested %>% 
  ggplot(aes(x = factor(obs_harvest_number), 
             y = n, fill = factor(obs_harvest_number))) +
  geom_col(col = 'black') +
  theme_xiaofei() +
  theme(legend.position="top") +
  theme(
    axis.text.x = element_text(size = 5, vjust = 1, angle = 65))+
  labs(x = "Harvest_plant_number", y = "Freq", fill = "Harvest_plant_number") +
  facet_wrap(~ use_trial_name)

ggsave(paste("images\\bar", trial_interest, Sys.Date(), ".png", sep = "_"),
       plot = plants_to_harvest, units = "in", dpi = 300, width = 9, height = 6)
```

## Compute germination and yield
Performs complex calculations on the dataset to derive new variables related to germination and yield

```{r}

plants_harvested$use_trial_name %>% unique()
  
trial_standard_new <- trial_standard %>%
  mutate(obs_harvest_number_plan =
           case_when(str_detect(use_trial_name, "202224") ~ 6,
                     str_detect(use_trial_name, "202255") ~ 6, 
                     str_detect(use_trial_name, "202326") ~ 6, 
                     str_detect(use_trial_name, "202327") ~ 6,
                     str_detect(use_trial_name, "202328") ~ 6,
                     str_detect(use_trial_name, "202329") ~ 6),
         obs_germination_perc = obs_germinated_number_plot/obs_planted_number_plot * 100,
         # 2) calculate area per plant
         area_plant = (use_plot_length*use_plot_width)/obs_planted_number_plot,
         # 3) calculate the yield_v4 obs_planted_number_plot
         obs_yield_ha_v2 = (((obs_root_weight_plot*10000)/(area_plant*obs_harvest_number_plan))/1000),
         obs_DM_yield_ha = obs_DM_gravity * obs_yield_ha / 100
        )
```

## Plot the yield_v2 vs yield upload by Lizbeth

```{r}
library(plotly)

p1 <- trial_standard_new %>% ggplot() +
  geom_point(aes(x = obs_yield_ha, y = obs_yield_ha_v2, color = use_plot_number), show.legend = F) +
  facet_wrap(~use_trial_name) +
  theme_xiaofei()

ggplotly(p1)
detach("package:plotly", unload = TRUE)
```


## Is numeric all traits?
Checks if all traits in the dataset are numeric, ensuring data type consistency for analysis.


```{r}
is_numeric(trial_data = trial_standard_new)
```

## Get the tidy data
Prepares the final tidy dataset for analysis by selecting and renaming variables, exemplifying data tidying practices.

```{r}
meta_info = names(trial_standard_new)[str_detect(names(trial_standard_new), "use_")]
meta_info = gsub("use_", "", meta_info)
meta_info
trial_tidy = trial_standard_new
names(trial_tidy)= gsub("use_", "", names(trial_standard_new))
# observations
trait_list = names(trial_tidy)[str_detect(names(trial_tidy), "obs_")]
trait_list = gsub("obs_", "", trait_list)
trait_list
names(trial_tidy)= gsub("obs_", "", names(trial_tidy))
trial_tidy = trial_tidy[c(meta_info, trait_list)]
```

# Boxplots and grouping boxplots
Generates boxplots for trait distributions across trials.

```{r warning=FALSE}
boxplot_traits(my_dat = trial_tidy, 
               folder = here::here("output//"), 
               trait_wanted = trait_wanted, 
               trial_interest = trial_interest)
```

## Grouping boxplot

```{r}

plot_bxp <- trial_tidy %>%
  pivot_longer(
    cols = all_of(trait_wanted),
    names_to = "var",
    values_to = "values"
  ) %>%
  filter(!var %in% c(
    "stake_plant", "planted_number_plot",
    "harvest_number", "root_weight_air",
    "root_weight_water", "harvest_number_plan",
    "yield_ha_v2", "root_rot_perc", "flesh_color_50inner1_8",
    "lodging1_3_6mon", "lodging1_3_4mon"
  )) %>%
  ggplot(aes(x = trial_name, y = values)) +
  geom_violin(fill = "gray") +
  geom_boxplot(width = 0.2) +
  labs(x = NULL, y = NULL, title = "") +
  theme_xiaofei() +
  theme(
    axis.text.x = element_text(size = 8, vjust = 1, angle = 75),
    axis.text.y = element_text(size = 8),
    plot.title = element_text(color = "black"),
    strip.text.x = element_text(
      size = 7, face = "bold.italic")
  ) +
  facet_wrap(~var,
    ncol = 7, scales = "free_y"
  ) 
print(plot_bxp)
ggsave(paste0("images\\boxplot_fixed_", trial_interest, Sys.Date(), ".png"),
    plot = plot_bxp, units = "in", dpi = 300, width = 12, height = 10
  )

```

## Save the tidy data for analysis.
Writes the cleaned and processed dataset to a CSV file for further analysis.

```{r eval=FALSE}
write.csv(trial_tidy, here::here("output", paste("01_", year_interest, trial_interest,
  "_tidy_data4analysis_", Sys.Date(), ".csv", sep = "")), row.names = FALSE)
```
